---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
#Loading and preprocessing the data

```{r, echo = TRUE}
activityData <- read.csv("C:\\Users\\e571237\\Desktop\\activity.csv")
```

#What is mean total number of steps taken per day?

##Calculate the total number of steps taken per day

```{r, echo = TRUE}
#splitting the data by date 

indDate <- split( x= activityData, f= activityData$date)

#creating a data frame that has the day no. and the no. of steps taken per day

df <- data.frame(Day = numeric(),
                 Total_Steps_taken_per_day= numeric(), 
                 stringsAsFactors=FALSE) 


for (i in 1:length(indDate))
{
  df <- rbind(df,sum(indDate[[i]]$steps))
}

df <- cbind(1:61, df)
names(df) <- c('Day','Total Steps taken per day')
```

##Make a histogram of the total no. of steps taken per day

```{r, echo = TRUE}

hist(df$`Total Steps taken per day`, main = "Total number of steps taken each day", xlab = "Steps", ylab = "Frequency")

```

##Calculate and report the mean and median of the total number of steps taken per day

```{r, echo = TRUE}
#created a data frame that has the day no, mean no of steps for that day and median no. of steps 
#for that day

meanSteps <- numeric(0)

for(i in 1:length(indDate))
{
  meanSteps <- append( meanSteps , mean(indDate[[i]]$steps, na.rm = TRUE))
}

medianSteps <- numeric(0)

for(i in 1:length(indDate))
{
  medianSteps <- append(medianSteps, median(indDate[[i]]$steps, na.rm = TRUE))
}

df <- cbind(df, meanSteps, medianSteps)
```

#What is the average daily activity pattern?

##Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo = TRUE}
indInt <- split( x= activityData, f= activityData$interval)

meanStepsByInt <- numeric(0)

for(i in 1:length(indInt))
{
  meanStepsByInt <- append( meanStepsByInt , mean(indInt[[i]]$steps, na.rm = TRUE))
}

intervalList <- unique(activityData$interval)

#time series plot of the average daily pattern of steps for 5-minute intervals
plot(x= intervalList, y = meanStepsByInt, type = "l", main = "Average daily pattern of steps", xlab = "Interval", ylab = "Average no. of steps")
```

##Which 5-minute interval on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo = TRUE}

index <- match(max(meanStepsByInt), meanStepsByInt)

intervalList[index]

```

#Imputing missing values

##1. Calculate and report the total number of missing values in the dataset(the total no. of rows with NAs)

```{r, echo = TRUE}

sum(is.na(activityData$steps))

###calculation part:
unique(is.na(activityData$interval)) 

unique(is.na(activityData$date)) 

#the above two statements returned false which means that there are no missing values for date or interval. Therefore, the missing values are present only in the steps column


###reporting part:

NaDF<-activityData[is.na(activityData$steps), ]     # this subset contains all the NA data 
unique(NaDF$date) #lists down all the dates that have all NA values
```

##Devise a strategy for filling in all of the missing values in the dataset.

IMPUTATION STRATEGY: Filling NAs with 0s
There are no NA values for "interval" or "date". So we need to fill in NA in "steps".

##Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r, echo = TRUE}

##filling NA values with 0s and creating a new dataset that is equal to the original dataset but with missing data filled in

activityData2 <- activityData
activityData2[is.na(activityData2)] <- 0

```

##Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day

```{r, echo = TRUE}
df2 <- data.frame(Day = numeric(),
                 Total_Steps_taken_per_day= numeric(), 
                 stringsAsFactors=FALSE) 

indDate2 <- split( x= activityData2, f= activityData2$date)

for (i in 1:length(indDate2))
{
  df2 <- rbind(df2,sum(indDate2[[i]]$steps))
}

df2 <- cbind(1:length(indDate2), df2)
names(df2) <- c('Day','Total Steps taken per day')

```

## Make a histogram of the total no. of steps taken per day

```{r, echo = TRUE}

hist(df2$`Total Steps taken per day`, main = "Total no of steps", xlab = "Steps", ylab = "Frequency")

###Filling NAs with 0s has had no effect on the total daily no. of steps
```

#Are there differences in activity patterns between weekdays and weekends?

##Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or a weekend day.
```{r, echo = TRUE}
library(timeDate)

LogicWeekday <- isWeekday(as.Date(activityData2$date), wday = 1:5)

Day <- character(0)

for ( i in 1:length(LogicWeekday))
{
  if (LogicWeekday[i] == TRUE)
    Day <- append(Day, "Weekday")
  else
    Day <- append(Day, "Weekend")
}

activityData3 <- cbind(activityData2, Day)
        #created a dataframe that has a column showing if a day is a weekday or a weekend
```


##Make a panel plot containing a time series plot (type = "l") of the 5-minute interval (x-axis) and the average no.of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r, echo = TRUE}

weekdayData <- subset(activityData3, Day == "Weekday")
weekendData <- subset(activityData3, Day == "Weekend")

indInt_wd <- split( x= weekdayData, f= weekdayData$interval)

meanStepsByInt_wd <- numeric(0)

for(i in 1:length(indInt_wd))
{
  meanStepsByInt_wd <- append( meanStepsByInt_wd , mean(indInt_wd[[i]]$steps, na.rm = TRUE))
}

intervalList_wd <- unique(weekdayData$interval)


indInt_wnd <- split( x= weekendData, f= weekendData$interval)

meanStepsByInt_wnd <- numeric(0)

for(i in 1:length(indInt_wnd))
{
  meanStepsByInt_wnd <- append( meanStepsByInt_wnd , mean(indInt_wnd[[i]]$steps, na.rm = TRUE))
}

intervalList_wnd <- unique(weekendData$interval)

#Time series plot of the 5-minute interval and the average no. of steps taken, averaged across all weekday days 
#or weekend days

library("graphics")

par(mfrow=c(2,1))

plot(x= intervalList_wd, y = meanStepsByInt_wd, type = "l", main = "Weekday", xlab = "Interval", ylab = "Average no. of steps")
plot(x= intervalList_wnd, y = meanStepsByInt_wnd, type = "l", main = "Weekend", xlab = "Interval", ylab = "Average no. of steps")
```


